---
title: "bspline-introduction tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bspline-introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(repos = c(CRAN = "https://cloud.r-project.org"))
```

# Introduction

This vignette serves as a tutorial for how to use the cubicSpline package

```{r setup}
library(cubicSpline)
```

To use the main function 'my_bspline' to generate the bspline coefficients

```{r}
set.seed(625) # For reproducibility

# Generate sample variable x and its response y
x <- seq(0, 10, length.out = 20)
y <- sin(x) + 2*cos(x)

my_bspline(x, y)
```

To plot the obtained spline against the original points using both 'my_bspline' and 'bspline_basis':

```{r}
# Function for visualization
visualize_my_bspline <- function(x, y) {
  coef <- my_bspline(x, y)   # Get coefficients using original function
  
  # Params set up
  m <- length(x)
  order <- 4
  internal_knots <- seq(min(x), max(x), length.out = m - order + 2)
  knots <- c(rep(min(x), order-1), internal_knots, rep(max(x), order-1))

  # Create grid for smooth plotting
  x_grid <- seq(min(x), max(x), length.out = 200)

  # Create matrix to store basis function values
  b_grid <- matrix(0, nrow = length(x_grid), ncol = m)

  # Calculate basis functions
  for (i in 1:length(x_grid)) {
    for (j in 1:m) {
      b_grid[i,j] <- bspline_basis(x_grid[i], j, order-1, knots)
    }
  }

  # Calculate weighted basis functions by multiplies each column of b_grid by the corresponding element in coef.
  b_weighted <- b_grid * rep(coef, each = nrow(b_grid))
  y_grid <- rowSums(b_weighted) # Calculate final sum

  # Plots:
  par(mfrow = c(3,1), mar = c(4,4,2,1))

  # The plot for original basis functions
  matplot(x_grid, 
          b_grid, 
          type = "l", 
          main = "Original basis Functions",
          xlab = "x", ylab = "Value")

  # The plot for weighted basis functions
  matplot(x_grid, 
          b_weighted, 
          type = "l", 
          main = "Weighted basis Functions (after multiplying by coefficients)",
          xlab = "x", ylab = "Value")

  # The plot for final sum and data points
  plot(x_grid, 
       y_grid, 
       type = "l", 
       main = "Final b-spline (sum of weighted basis functions)",
       xlab = "x", ylab = "Value", col = "pink")
  points(x, y, pch = 16, col = "purple")
  par(mfrow = c(1,1))
}

visualize_my_bspline(x,y)
```

# Compare against fda package

## The accuracy comparison

```{r}
# Install and load package
install.packages("fda")
library(fda)

# B-spline coefficients match fda package for cos function
x <- seq(-1, 1, length.out = 15)
y <- cos(x)
my_coef <- my_bspline(x, y)

# Obtain the coefficients result from fda package
degree <- 3
n_basis <- length(x)
basis <- create.bspline.basis(
  rangeval = c(min(x), max(x)),
  nbasis = n_basis,
  norder = degree + 1
)
basis_matrix <- eval.basis(x, basis)
fit_fda <- lm(y ~ basis_matrix - 1)
fda_coef <- as.numeric(fit_fda$coefficients) #Transform coefficients to include just the value

all.equal(my_coef, fda_coef) # Test equality of coefficients

```

## The efficiency comparison

Benchmarking:

```{r}
install.packages("microbenchmark")
install.packages("dplyr")
install.packages("ggplot2")

library(microbenchmark)
library(dplyr)
library(ggplot2)


# Create separate functions of the fda methods for easier benchmarking
fda_spline <- function() {
  degree <- 3
  n_basis <- length(x)
  basis <- create.bspline.basis(
    rangeval = c(min(x), max(x)),
    nbasis = n_basis,
    norder = degree + 1
  )
  basis_matrix <- eval.basis(x, basis)
  fit_fda <- lm(y ~ basis_matrix - 1)
  fda_coef <- as.numeric(fit_fda$coefficients)
  return(fda_coef)
}

# Run the benchmark
benchmark_results <- microbenchmark(my_spline = my_bspline(x, y), fda_spline = fda_spline(),times = 100)

# Print the benmark results
print(benchmark_results)
```

Visualization of benchmark result: 

```{r}

# Convert to data frame and adjust time to be in milliseconds
results <- as.data.frame(benchmark_results) %>% mutate(time_ms = time / 1e6) 

# Create the plot
ggplot(results, aes(x = time_ms, y = expr)) +
  geom_jitter(aes(color = expr), height = 0.2, alpha = 0.5) +
  scale_x_log10(
    breaks = c(0.01, 0.1, 1, 10, 100),
    labels = c("10µs", "100µs", "1ms", "10ms", "100ms")
  ) +
  labs(
    x = "time",
    y = "expression",
    color = "method"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "right"
  )
```



```{r}
# Generate example data
set.seed(625)
x <- seq(0, 1, length.out = 10)
y <- sin(2*pi*x) + rnorm(10, 0, 0.1)

# Get coefficients
coef <- my_bspline(x, y)

# Predict at new points
x_new <- seq(0, 1, length.out = 100)
y_pred <- numeric(length(x_new))

# Calculate predictions using existing basis functions
order <- 4
internal_knots <- seq(min(x), max(x), length.out = length(x) - order + 2)
knots <- c(rep(min(x), order-1), internal_knots, rep(max(x), order-1))
for(i in 1:length(x_new)) {
  y_pred[i] <- sum(coef * sapply(1:length(coef), function(j) 
    bspline_basis(x_new[i], j, order-1, knots)))
}
```

